preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-15 Manuscript/manuscript.Rmd
Results polishing

In our ML framework boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 cell line-specific functional genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 2). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial (Distance) relationships (see Methods). A ($log_2 + 1$)-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1).

@@

![**Distance-type predictors combined with random under-sampling at 5 kb resolution genomic bins yielded optimal TAD boundary region predictive performance for GM12878 Arrowhead TAD boundaries.** Averaged balanced accuracies are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate 1 standard deviation away from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png)

@@

## Predictive importances confirmed biological functionality of CTCF, RAD21, SMC3, and ZNF143 at TAD boundaries

We sought to further optimize our boundary region prediction models by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each set of chromosomes. To determine transcription factors most important for boundary prediction, we clustered the predictive importances (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were consistently predictive of TAD and loop boundaries (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line when trained on another (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic elements from GM12878 resulted in an average AUC=0.845 (Figure 4D). However when training on GM12878 and testing on K562, our model's average AUC only marginally decreased to 0.803. Likewise, training and testing on Peakachu resulted in an average AUC=0.885 and decreased only slightly to 0.868 when trained on GM12878 and tested on K562. Results were consistent when training on K562 and testing on GM12878 (Supplementary Figure 4D). These results suggest that a classifier trained on the data from one cell line performs well when using data from another cell line.

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space was built on a base-resolution level, and the model trained on a bin-level space was used to predict the probability of each base being a boundary. Our method, referred to as *preciseTAD*, uses density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location at base resolution (see Methods).

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7,846 TAD and 15,838 chromatin loop boundaries on GM12878, as well as 8,883 TAD and 11,016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 3). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that often do not correspond to intra-chromosomal contacts featured on a contact matrix. Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 3). This can be contributed to the fact that *preciseTAD* leverages four known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting Peakachu which considers only CTCF sites as true positives. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant, in contrast to boundaries detected from Hi-C data alone.

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 8). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results on Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar protein complexes.

>>>>

Boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 cell-line-specific genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 2). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial ($log_2 + 1$ Distance) relationships (see Methods, Supplementary Figure 1).

@@

![**Distance-type predictors combined with random under-sampling at 5 kb resolution genomic bins yielded optimal TAD boundary region predictive performance for GM12878 Arrowhead TAD boundaries.** Averaged balanced accuracies are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate standard deviation away from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png)

@@

## Predictive importances confirmed biological role of CTCF, RAD21, SMC3, and ZNF143 for boundary formation

We sought to further optimize our boundary region prediction models by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each set of chromosomes. To determine transcription factors most important for boundary prediction, we clustered the predictive importances (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were consistently predictive of TAD and loop boundaries (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line when trained on another (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic elements from GM12878 resulted in an average AUC=0.845 (Figure 4D). When training on GM12878 and testing on K562, our model's average AUC only marginally decreased to 0.803. Likewise, training and testing on Peakachu resulted in an average AUC=0.885 and decreased only slightly to 0.868 when trained on GM12878 and tested on K562. Results were consistent when training on K562 and testing on GM12878 (Supplementary Figure 4D). These results suggest that a classifier trained on the data from one cell line performs well when using data from another cell line.

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict boundaries at base resolution. Intuitively, instead of bin-level annotations, the predictor-response space was built on a base-level, that is, each base was associated with the corresponding genomic annotation. The model trained on a bin-level space was then applied on a base-level space to predict the probability of each base being a boundary. Our method, referred to as *preciseTAD*, uses density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location at base resolution (see Methods).

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7,846 TAD and 15,838 chromatin loop boundaries on GM12878, as well as 8,883 TAD and 11,016 chromatin loop boundaries on K562 [??? Need them in Supplementary Table 4, Excel, BED format, sheet 1 - GM12878 PTBR, sheet 2 - GM12878 PTBP, sheet 3,4 - K562. Supp tables 1, 2, 3 are OK, renumber subsequent. Write legend]. We reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 3). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that, when visualized, often do not correspond to intra-chromosomal contacts featured on a contact matrix. Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 3). This can be attributed to the fact that *preciseTAD* leverages four known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting Peakachu which considers only CTCF sites to mark the predicted boundaries as true positives. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant, in contrast to boundaries detected from Hi-C data alone.

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 8). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results on Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop domain boundaries may be driven by similar molecular mechanisms.
